Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2021-09-13 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
auto number tables

Elemental concentrations such as $N_{area}$ and $P_{area}$, contribute to high rates of leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table 1, @scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014], and may be optimized throughout the vertical canopy gradient [@buckleyOptimalCarbonPartitioning2021]. Within-canopy leaf nitrogen distribution can be regulated to prevent photoinhibition in shade leaves during sunflecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018]. However, @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between total foliar concentrations of N and P and foliar thermal tolerance.

Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table 1). 
Carotenoids and xanthophyll cycle pigments -- violaxanthin, antheraxanthin and zeaxanthin (VAZ) -- can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996]. Heat-sensitive emissions of isoprene gas from tree leaves enhance photosynthetic thermal tolerance by regulating antioxidant defences and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021].

@@

Maximum $g_s$ increases with light, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table 2). However, because of the high VPD and irradiance experienced by sun leaves,  midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests.  In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021a].

@@

**Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies [Table 2; @niinemetsPhotosynthesisResourceDistribution2007 ; @kenzoHeightrelatedChangesLeaf2015 ; @slotPhotosyntheticHeatTolerance2019; @chenLeafEconomicsSpectrum2020].**
This is primarily driven by the greater light available to sun leaves, which also have traits enabling higher photosynthetic rates under high light and sufficient nutrient and water conditions (Table 1, Table 2).

@@

Leaf thermal tolerances are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a]. In Australia, along a vertical gradient within the canopy, *Acacia papyrocarpa* (Benth.) showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing canopy than other canopy positions, correlating with low wind-speed, greater radiation and $T_{air}$, and lower rates of heat dissapation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019]. Additionally, thermal tolerances are explained more by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. $T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time, whereas $T_{50}$ was greater among species with greater LMA [@slotLeafHeatTolerance2021]. Across species sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. However, considering sun and shade leaf differences within canopy (Table 1), $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019]. The small difference in $T_{50}$ coupled with large differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].

@@

**Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [Table 2, @chenLeafEconomicsSpectrum2020], but its temperature sensitivity ($Q_{10}$) shows no definite trend along the vertical gradient [@bolstad_foliar_1999; @weerasingheCanopyPositionAffects2014].** Specifically, the $Q_{10}$ of respiration can be constant within vertical profiles and in seedling sun vs. shade leaves [@carterExperimentalWarmingTropical2021a; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002]. This may vary based on forest type [**REF**], acclimation to high temperature that decreases $Q_{10}$ [@slotGeneralPatternsAcclimation2015], or leaf traits and leaf age, as in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015]. Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising $T_{air}$ [@carterExperimentalWarmingTropical2021a].

@@

**Vertical gradients affect tree metabolism and growth.**
Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured as diameter at breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, and height, all of which scale with DBH [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015].
Specifically, increases are linked to increased leaf area and in part to an increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific transpiration and photosynthesis (Table 2).
The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs). 
Within a stand, woody growth rates , measured as radial stem growth or biomass increment,  generally increase with tree size [e.g., @muller-landauTestingMetabolicEcology2006, Stephenson et al. 2014]. NSC reserves also increase with tree size, such that larger trees have relatively more buffer against biological and biophysical stressors [@niinemetsResponsesForestTrees2010a].

**Vertical gradients also affect the climate sensitivity of metabolism and growth.** 
Stomatal conductance tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and VPD, low RH) even before soil water becomes limited in forests not experiencing drought [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016]. Therefore, periods of high atmospheric demand--be these on time scales of hours, days, or seasons--tend to cause greater reductions in whole-tree transpiration and photosynthesis in tall trees that occupy canopy positions in relatively dense-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021]. This is consistent with the observation that $g_s$-limitation with $T_{air}$ increases with height in the canopy (Table 2). More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004]-- often combined with greater effective rooting depth--may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and lower capacity to adjust traits related to carbon metabolism to drought [@bartholomewSmallTropicalForest2020]. This results in decreasing carbon isotope discrimination, indicative of an increasingly conservative hydraulic strategy, with increasing tree height [Table 1, @mcdowellRelationshipTreeHeight2011].

@@

![**Figure `r fig_model`.** Lidar and thermal remote sensing data can be used to measure vertical canopy structure and plant thermal signatures at increasingly large scales. (a-b) Integrating in situ and remote sensing measurements will support our ability to scale understanding of vertical gradients within and across ecosystems. (c) In turn, these efforts can be used to directly inform ecosystem model development and parameterization. Vertically resolved in situ data could include any variable listed in Table 1 or 2.](schematics/fig_remote_sensing.png)

>>>>

Elemental concentrations such as $N_{area}$ and $P_{area}$ contribute to high rates of leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table `r table_leaf_traits`, @scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014], and may be optimized throughout the vertical canopy gradient [@buckleyOptimalCarbonPartitioning2021]. 
Within-canopy leaf nitrogen distribution can be regulated to prevent photoinhibition in shade leaves during sunflecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018]. 
However, @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between total foliar concentrations of N and P and foliar thermal tolerance.

Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
Carotenoids and xanthophyll cycle pigments -- violaxanthin, antheraxanthin and zeaxanthin (VAZ) -- can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996]. Heat-sensitive emissions of isoprene gas from tree leaves enhance photosynthetic thermal tolerance by regulating antioxidant defences and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021].

@@

Maximum $g_s$ increases with light, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`). However, because of the high VPD and irradiance experienced by sun leaves,  midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests.  In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021a].

@@

**Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies [Table `r table_leaf_metabolism`; @niinemetsPhotosynthesisResourceDistribution2007 ; @kenzoHeightrelatedChangesLeaf2015 ; @slotPhotosyntheticHeatTolerance2019; @chenLeafEconomicsSpectrum2020].**
This is primarily driven by the greater light available to sun leaves, which also have traits enabling higher photosynthetic rates under high light and sufficient nutrient and water conditions (Tables `r table_leaf_traits`, `r table_leaf_metabolism`).

@@

Leaf thermal tolerances are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a]. In Australia, along a vertical gradient within the canopy, *Acacia papyrocarpa* (Benth.) showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing canopy than other canopy positions, correlating with low wind-speed, greater radiation and $T_{air}$, and lower rates of heat dissapation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019]. Additionally, thermal tolerances are explained more by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. $T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time, whereas $T_{50}$ was greater among species with greater LMA [@slotLeafHeatTolerance2021]. Across species sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. However, considering sun and shade leaf differences within canopy (Table `r table_leaf_traits`), $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019]. The small difference in $T_{50}$ coupled with large differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].

@@

**Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [Table `r table_leaf_metabolism`, @chenLeafEconomicsSpectrum2020], but its temperature sensitivity ($Q_{10}$) shows no definite trend along the vertical gradient [@bolstad_foliar_1999; @weerasingheCanopyPositionAffects2014].** Specifically, the $Q_{10}$ of respiration can be constant within vertical profiles and in seedling sun vs. shade leaves [@carterExperimentalWarmingTropical2021a; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002]. This may vary based on forest type [**REF**], acclimation to high temperature that decreases $Q_{10}$ [@slotGeneralPatternsAcclimation2015], or leaf traits and leaf age, as in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015]. Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising $T_{air}$ [@carterExperimentalWarmingTropical2021a].

@@

**Vertical gradients affect tree metabolism and growth.**
Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured as diameter at breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, and height, all of which scale with DBH [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015].
Specifically, increases are linked to increased leaf area and in part to an increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific transpiration and photosynthesis (Table `r table_leaf_metabolism`).
The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs). 
Within a stand, woody growth rates , measured as radial stem growth or biomass increment,  generally increase with tree size [e.g., @muller-landauTestingMetabolicEcology2006, Stephenson et al. 2014]. NSC reserves also increase with tree size, such that larger trees have relatively more buffer against biological and biophysical stressors [@niinemetsResponsesForestTrees2010a].

**Vertical gradients also affect the climate sensitivity of metabolism and growth.** 
Stomatal conductance tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and VPD, low RH) even before soil water becomes limited in forests not experiencing drought [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016]. Therefore, periods of high atmospheric demand--be these on time scales of hours, days, or seasons--tend to cause greater reductions in whole-tree transpiration and photosynthesis in tall trees that occupy canopy positions in relatively dense-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021]. This is consistent with the observation that $g_s$-limitation with $T_{air}$ increases with height in the canopy (Table `r table_leaf_metabolism`). More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004]-- often combined with greater effective rooting depth--may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and lower capacity to adjust traits related to carbon metabolism to drought [@bartholomewSmallTropicalForest2020]. This results in decreasing carbon isotope discrimination, indicative of an increasingly conservative hydraulic strategy, with increasing tree height [Table `r table_leaf_traits`, @mcdowellRelationshipTreeHeight2011].

@@

![**Figure `r fig_model`.** Lidar and thermal remote sensing data can be used to measure vertical canopy structure and plant thermal signatures at increasingly large scales. (a-b) Integrating in situ and remote sensing measurements will support our ability to scale understanding of vertical gradients within and across ecosystems. (c) In turn, these efforts can be used to directly inform ecosystem model development and parameterization. Vertically resolved in situ data could include any variable listed in Table `r table_leaf_traits` or 2.](schematics/fig_remote_sensing.png)
